---
title: "ggmapping"
layout: post
comments: true
---

{% raw %}

# ggmapping

R has several packages for making maps, including the packages `maps` and `ggmap`. The latter uses smooth aesthetics of `ggplot` to make really good-looking maps. In `ggmaps`, there are several options for map tile types or static map backgrounds, including Google and Stamen maps. Check out [this ggmap cheat sheet](https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/ggmap/ggmapCheatsheet.pdf) to get a look at the different mapping backgrounds available. Stamen maps take a little while to download at high resolution, but yield a nice final product. One downside of using `ggmap` is that global maps aren't available through Google or Stamen maps. 

Here, you can download a Stamen map of the native monk parakeet range, then overlay GBIF sightings by country. 

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(ggmap)
library(ggplot2)
library(maptools)
library(scales) 
library(sp)
library(rgdal)
```

```{r, echo = TRUE, eval = FALSE}
library(ggmap)
library(ggplot2)
library(maptools)  #for shapefiles
library(scales)  #for transparency
library(sp)
library(rgdal)
```

```{r, eval = TRUE, echo = FALSE}
mp <- read.csv("/Users/gracesmithvidaurre/Desktop/R/gsmithvi.github.io/_data/mymon.csv", header = TRUE, row.names = 1)
```

```{r, eval = FALSE, echo = TRUE}
mp <- read.csv(paste("Myiopsitta_monachus_filtered_GBIF_", Sys.Date(), ".csv", sep = ""), header = TRUE)
```

The GBIF data needs to be filtered for South American sightings
```{r, echo = TRUE, eval = TRUE}
mp1 <- mp[mp$decimalLatitude <= 0 & mp$decimalLongitude <= 0, ]
```

How many countries are in this filtered dataset?
```{r, echo = TRUE, eval = TRUE}
table(mp1$countryCode)
```

The `countryCode` variable is a factor with empty levels, 
these empties need to be dropped before mapping
```{r, echo = TRUE, eval = TRUE}
mp1$countryCode <- droplevels(mp1$countryCode)
table(mp1$countryCode)
```

Download the Stamen map, you need Internet for this. A higher zoom argument makes the download time much longer, but gives you better resolution of waterways and other natural boundaries
```{r, echo = TRUE, eval = FALSE}
bbox <- c(left = -80, bottom = -45, right = -40, top = 2)
smap <- get_stamenmap(bbox = bbox, zoom = 8,
                      maptype = c("toner-background"))
```

Read in species distribution polygon and initialize data for mapping
```{r, echo = TRUE, eval = FALSE}

mp.distr <- readOGR("./_data", "Myiopsitta_monachus_1608_NS")

dats <- data.frame(lon = mp1$decimalLongitude, lat = mp1$decimalLatitude, 
                   country = mp1$countryCode)

fill.cols <- diverge_hcl(length(table(dats$co)), alpha = 0.6)

cols <- rep("black", length(fill.cols))
```

Overlay GBIF sightings and the species distribution onto the Stamen map.
```{r, echo = TRUE, eval = FALSE}
tiff("mymon-south-america.tiff", res = 300, units = "cm", width = 20,
     height = 20)

ggmap(smap) + 
  
  geom_polygon(aes(x = long, y = lat, group = group), color = topo.colors(12)[5], fill = topo.colors(12, alpha = 0.6)[7], size = 1, data = mp.distr) +
  
  geom_point(aes(x = lon, y = lat, color = country, fill = country, size = country, shape = country), data = dats) + 
  
  scale_fill_manual(values = fill.cols) +
  
  scale_colour_manual(values = cols) + 
  
  scale_shape_manual(values = rep(21, length(fill.cols))) + 
  
  scale_size_manual(values = rep(4, length(fill.cols))) 

dev.off()

```

<img src="/images/mymon-south-america.tiff" width="550" height ="600" />

{% endraw %}